Division and Slope Factorization of p-Adic Polynomials

### 2016-04-30 polynomials/slopefact.tex
Merge David's corrections in slopefact.tex

Polynomial factorization over local fields has seen a great deal of progress recently
\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
following an algorithm of Montes.  Slope factorization provides a subroutine
in such algorithms \cite[Section~2]{pauli:10a}.  For the most difficult inputs, it is not the dominant
contributor to the runtime of the algorithm, but in some circumstances it will be.
We underline moreover that the methods introduced in this paper extend 
partially to the noncommutative setting and appear this way as an
essential building block in several decomposition algorithms of 
$p$-adic Galois representations and $p$-adic differential 
equations~\cite{caruso:16}.

@@

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_n X^n \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \cite[Section~1.6]{dwork-geratto-sullivan:Gfunctions} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \nftimes \NF(Q)
\end{array}$$
where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
There exist classical algorithms for computing these two operations
whose complexity is quasi-linear with respect to the degree.

@@

In the sequel, it will be convenient to use a different normalization on 
$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
always assume that $P$ is monic in the sense that its leading
coefficient is \emph{exactly}~$1$; the precision datum of $P$ then 
only concerns the coefficients up to degree $n{-}1$. Similarly, noting
that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.

@@

\begin{rem}
The precision needed at line 6 is of course governed by the computation
performed at line 7. Note that it can be either computed \emph{a
priori} by using Proposition~\ref{prop:NewtonprecEuclide} or dynamically 
by using relaxed algorithms from 
\cite{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
\NF(P))$.
\end{rem}

@@

\noindent
Obviously \textbf{(H)} gives an answer to the first question we have 
raised above: the precision of the couple $A^{(1)}$ is the lattice 
$H_{A^{(1)}}$ defined as projection on the first component of 
$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
shown by the next proposition.

\begin{prop}
\label{prop:precA1}
The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at some point $P$ is the linear mapping
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}

@@

As a conclusion, the algorithm we propose consists in computing the 
Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of the
algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
until we find a couple $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
couple has been found, one may safely output $\tilde A^{(1)} + 
O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under
\textbf{(H)}.
The resulting algorithm has quasi-optimal running 
time and optimal stability.

>>>>

Polynomial factorization over local fields has seen a great deal of progress recently
\citelist{\cite{pauli:10a} \cite{guardia-nart-pauli:12a} \cite{guardia-montes-nart:08a} \cite{montes:99a}}
following an algorithm of Montes.  Slope factorization provides a subroutine
in such algorithms \cite[Section~2]{pauli:10a}.  Note that our algorithm lifts
a \emph{slope factor} corresponding to a segment of the Newton polygon,
rather than the irreducible factor found by single factor lifting \cite{guardia-nart-pauli:12a}.
The slope factors are easier to find; as a consequence our iteration is less involved.
We also remark that the methods introduced in this paper extend 
partially to the non-commutative setting and appear this way as an
essential building block in several decomposition algorithms of 
$p$-adic Galois representations and $p$-adic differential 
equations~\cite{caruso:16}.

@@

We recall that one can attach a Newton function to each polynomial.
If $P(X) = \sum_{i=0}^n a_i X^i \in K_n[X]$, we define its Newton
polygon $\NP(P)$ as the convex hull of the points $(i,\val(a_i))$ 
($1 \leq i \leq n$) together with the point at infinity $(0,+\infty)$
and then its Newton function $\NF(P) : [0,n] \to \R$ as the unique
function whose epigraph is $\NP(P)$. It is well known \cite[Section~1.6]{dwork-geratto-sullivan:Gfunctions} that:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NP(P+Q) & \subset & \text{Conv}\big(\NP(P) \cup \NP(Q)\big) \smallskip \\
\NP(PQ) & = & \NP(P) + \NP(Q)
\end{array}$$
where $\text{Conv}$ denotes the convex hull and the plus sign stands 
for the Minkowski sum. This translates to:
$$\begin{array}{r@{\,\,}c@{\,\,}l}
\NF(P+Q) \geq \NF(P) \nfplus \NF(Q) \smallskip \\
\NF(PQ) = \NF(P) \nftimes \NF(Q)
\end{array}$$
where the operations $\nfplus$ and $\nftimes$ are defined accordingly.
There exist classical algorithms for computing these two operations
whose complexity is quasi-linear with respect to the degree.

@@

In the sequel, it will be convenient to use a different normalization on 
$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We also assume
the leading coefficient of $P$ is \emph{exactly}~$1$; the precision datum on $P$ then 
only concerns the coefficients up to degree $n{-}1$. Similarly,
$A^{(1)}$ and $B^{(1)}$ only carry
precision up to degree $d{-}1$ and $n{-}d{-}1$ respectively.

@@

\begin{rem}
The precision needed at line 6 is of course governed by the computation
performed at line 7. Note that it can be either computed \emph{a
priori} by using Proposition~\ref{prop:NewtonprecEuclide} or dynamically 
by using relaxed algorithms from 
\cites{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
\NF(P))$.
\end{rem}

@@

\noindent
Obviously \textbf{(H)} gives an answer to the first question we have 
raised above: the precision of $A^{(1)}$ is the lattice 
$H_{A^{(1)}}$ defined as projection on the first component of 
$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
shown by the next proposition.

\begin{prop}
\label{prop:precA1}
The map $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at some point $P$ is the linear mapping
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}

@@

As a conclusion, the algorithm we propose consists in computing the 
Newton sequences $(A_i)$ and $(V_i)$ (following the strategy of
Algorithm \texttt{slope\_factorisation\_Newton} regarding to precision)
until we find a pair $(\tilde A^{(1)}, \tilde V^{(1)})$ satisfying 
the requirements (i) and (ii) of Proposition~\ref{prop:stop}. Once this 
pair has been found, one may safely output $\tilde A^{(1)} + 
O\big(\big(\tilde V^{(1)} \cdot H_P\big) \mod \tilde A^{(1)}\big)$ under
\textbf{(H)}.
The resulting algorithm has quasi-optimal running 
time and optimal stability.
